---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
mainfont: Roboto
monofont: Consolas
author: "Giang Truong"
output: pdf_document
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```

```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth))
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n))
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})
```

# Task 1

## Load required libraries and datasets

Note that you will need to install these libraries if you have never used these before.

```{r  Load Libraries, results = 'hide'}

#### Load required libraries
library(data.table)
library(ggplot2)
library(readr)
library(readxl)
library(dplyr)
#### Point the filePath to where you have downloaded the datasets to and
#### assign the data files to data.tables
setwd("C:/Giang/Studying/Virtual Internship- Quantium")
transactionData <- read_excel("QVI_transaction_data.xlsx")
customerData <- read.csv("QVI_purchase_behaviour.csv")
```

## Exploratory data analysis

The first step in any analysis is to first understand the data. Let's take a look at each of the datasets provided. \### Examining transaction data We can use `str()` to look at the format of each column and see a sample of the data. As we have read in the dataset as a `data.table` object, we can also run `transactionData` in the console to see a sample of the data or use `head(transactionData)` to look at the first 10 rows. Let's check if columns we would expect to be numeric are in numeric form and date columns are in date format.

```{r Examining transaction data}
#### Examine transaction data
str(transactionData)
transactionData <- data.table(transactionData)
```

We can see that the date column is in an integer format. Let's change this to a date format.

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

We should check that we are looking at the right products by examining PROD_NAME.

```{r Summary of PROD_NAME}
#### Examine PROD_NAME
head(transactionData$PROD_NAME)
str(transactionData$PROD_NAME)
summary(transactionData$PROD_NAME)
```

Looks like we are definitely looking at potato chips but how can we check that these are all chips? We can do some basic text analysis by summarising the individual words in the product name.

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips
productWords <- data.table(unlist(strsplit(unique(transactionData$PROD_NAME), " ")))
setnames(productWords, 'words')
```

As we are only interested in words that will tell us if the product is chips or not, let's remove all words with digits and special characters such as '&' from our set of product words. We can do this using `grepl()`.

```{r}
#### Removing digits
productWords <- productWords[grepl("\\d",words)== FALSE, ]
#### Removing special characters
productWords <- productWords[grepl("[:alpha:]",words), ]
#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency
productWords[, .N, words][order(N, decreasing = TRUE)]
```

There are salsa products in the dataset but we are only interested in the chips category, so let's remove these.

```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max values for each feature to see if there are any obvious outliers in the data and if there are any nulls in any of the columns (`NA's : number of nulls` will appear in the output if there are any nulls).

```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
summary(transactionData)
```

There are no nulls in the columns but product quantity appears to have an outlier which we should investigate further. Let's investigate further the case where 200 packets of chips are bought in one transaction.

```{r }
#### Filter the dataset to find the outlier
transactionData[PROD_QTY == 200, ]

```

There are two transactions where 200 packets of chips are bought in one transaction and both of these transactions were by the same customer.

```{r}
#### Let's see if the customer has had other transactions
transactionData[LYLTY_CARD_NBR == 226000, ]
```

It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#### Filter out the customer based on the loyalty card number
transactionData <- transactionData[LYLTY_CARD_NBR != 226000, ]
#### Re-examine transaction data
summary(transactionData)
```

That's better. Now, let's look at the number of transaction lines over time to see if there are any obvious data issues such as missing data.

```{r}
#### Count the number of transactions by date
transactionData[, .(Count = .N), by = .(DATE)]

```

There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.

```{r fig.align = "center"}
#### Create a sequence of dates and join this the count of transactions by date

total_dates <- seq(as.Date("2018-07-01"), as.Date("2019-06-30"), by = "day")

# Join the date sequence with the transaction data to get the transaction counts
transactions_by_day <- transactionData[, .(Transaction_count = .N), by = .(DATE)]
transactions_by_day <- merge(data.table(DATE = total_dates), transactions_by_day, all.x = TRUE)

#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#### Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = Transaction_count)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

We can see that there is an increase in purchases in December and a break in late December. Let's zoom in on this.

```{r fig.align = "center"}
#### Filter to December and look at individual days
ggplot(transactions_by_day[month(DATE)==12], aes(x = DATE, y = Transaction_count)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 day") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on Christmas day. Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.

```{r Create pack size}
#### Pack size
#### We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]
#### Always check your output
#### Let's check if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r }
#### Let's plot a histogram of PACK_SIZE since we know that it is a categorical variable and not a continuous variable even though it is numeric.
hist(transactionData$PACK_SIZE)

```

Pack sizes created look reasonable. Now to create brands, we can use the first word in PROD_NAME to work out the brand name...

```{r Create brand name}

#### Brands
# Create a new column called "brand"
transactionData$BRAND <- NA

# Use regular expressions to extract the first word from the PROD_NAME column and assign it to the brand column

transactionData$BRAND[grep("^\\w+", transactionData$PROD_NAME)] <- 
  gsub("(^\\w+).*", "\\1", transactionData$PROD_NAME[grep("^\\w+", transactionData$PROD_NAME)])

# Note: grep("^\\w+", transactionData$PROD_NAME) searches the PROD_NAME column for strings that begin with one or more word characters (letters, digits, or underscores).
# gsub("(^\\w+).*", "\\1", transactionData$PROD_NAME[grep("^\\w+", transactionData$PROD_NAME)]) extracts the first word from each matching string in the PROD_NAME column using the gsub function, which replaces the entire string with just the first word. The regular expression (^\\w+).* matches the beginning of the string (^), one or more word characters (\\w+), and any characters after that (.*). The replacement string \\1 inserts just the matched word.

# Print the first 10 rows of the updated transactionData table
head(transactionData, 10)

#### Checking brands
transactionData[, .N, BRAND][order(-N)]

```

Some of the brand names look like they are of the same brands - such as RED and RRD, which are both Red Rock Deli chips. Let's combine these together.

```{r Clean brand names}
#### Clean brand names
transactionData[BRAND == "RED", BRAND := "RRD"]
transactionData[BRAND == "Dorito", BRAND := "Doritos"]
transactionData[BRAND == "WW", BRAND := "Woolworths"]
transactionData[BRAND == "NCC", BRAND := "Natural"]
transactionData[BRAND == "Infzns", BRAND := "Infuzions"]
transactionData[BRAND == "Smith", BRAND := "Smiths"]
transactionData[BRAND == "Snbts", BRAND := "Sunbites"]
transactionData[BRAND == "Grain", BRAND := "GrnWves"]
#### Check again
transactionData[, .N, BRAND][order(BRAND)]
# Over to you! Check the results look reasonable.
```

### Examining customer data

Now that we are happy with the transaction dataset, let's have a look at the customer dataset.

```{r 1 Exploratory data analysis}
#### Examining customer data
str(customerData)
summary(customerData)
head(customerData,10)
```

```{r }
#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be sure that no duplicates were created. This is because we created `data` by setting `all.x = TRUE` (in other words, a left join) which means take all the rows in `transactionData` and find rows with matching values in shared columns and then joining the details in these rows to the `x` or the first mentioned table Let's also check if some customers were not matched on by checking for nulls.

```{r Check for missing customer details}
sum(is.na(data$LIFESTAGE))
sum(is.na(data$PREMIUM_CUSTOMER))
```

Great, there are no nulls! So all our customers in the transaction data has been accounted for in the customer dataset. Note that if you are continuing with Task 2, you may want to retain this dataset which you can write out as a csv

```{r Code to save dataset as a csv}
write.csv(data,"QVI_data.csv")
```

Data exploration is now complete!

### Data analysis on customer segments

Now that the data is ready for analysis, we can define some metrics of interest to the client:

-   Who spends the most on chips (total sales), describing customers by lifestage and how premium their general purchasing behaviour is

\- How many customers are in each segment - How many chips are bought per customer by segment

\- What's the average chip price by customer segment

We could also ask our data team for more information. Examples are:

\- The customer's total spend over the period and total spend for each transaction to understand what proportion of their grocery spend is on chips

\- Proportion of customers in each customer segment overall to compare against the mix of customers who purchase chips

Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and plotting the split by these segments to describe which customer segment contribute most to chip sales.

```{r fig.width = 10, fig.align = "center"}
#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER
total_sales <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarize(total_sales = sum(TOT_SALES))

# Print the results
print(total_sales)

#Visualizing total sales: 
ggplot(total_sales, aes(x = LIFESTAGE, y = total_sales, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Total Sales by LIFESTAGE and PREMIUM_CUSTOMER", x = "LIFESTAGE", y = "Sales", fill = "PREMIUM_CUSTOMER")

```

Sales are coming mainly from Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees Let's see if the higher sales are due to there being more customers who buy chips.

```{r fig.width = 10, fig.align = "center"}
#### Number of customers by LIFESTAGE and PREMIUM_CUSTOMER
customers <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarize(customers = uniqueN(LYLTY_CARD_NBR))
print(customers)

ggplot(customers, aes(x = LIFESTAGE, y = customers, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Number of customers by LIFESTAGE and PREMIUM_CUSTOMER", x = "LIFESTAGE", y = "Number of customers", fill = "PREMIUM_CUSTOMER")

```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy chips. This contributes to there being more sales to these customer segments but this is not a major driver for the Budget - Older families segment. Higher sales may also be driven by more units of chips being bought per customer. Let's have a look at this next.

```{r fig.width = 10, fig.align = "center"}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
avg_units <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarize(avg_units = sum(PROD_QTY)/uniqueN(LYLTY_CARD_NBR))
print(avg_units)

ggplot(avg_units, aes(x = LIFESTAGE, y = avg_units, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER", x = "LIFESTAGE", y = "Average Units", fill = "PREMIUM_CUSTOMER")
# Over to you! Calculate and plot the average number of units per customer by those two dimensions.
```

Older families and young families in general buy more chips per customer Let's also investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.

```{r fig.width = 10, fig.align = "center"}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
avg_price <- data %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarize(avg_price = sum(TOT_SALES)/sum(PROD_QTY))
print(avg_price)

ggplot(avg_price, aes(x = LIFESTAGE, y = avg_price, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Average price per units by LIFESTAGE and PREMIUM_CUSTOMER", x = "LIFESTAGE", y = "Average Units", fill = "PREMIUM_CUSTOMER")

```

Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than their own consumption. This is also supported by there being fewer premium midage and young singles and couples buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn't large, we can check if this difference is statistically different.

```{r}
#### Perform an independent t-test between mainstream vs premium 
#### and budget midage and
#### young singles and couples
data$price <- data$TOT_SALES/data$PROD_QTY

t.test(data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER == "Mainstream", price],data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER != "Mainstream", price], alternative = "greater" )

```

The t-test results in a p-value \< 2.2e-16 which is \< 0.05, i.e. the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples.

\## Deep dive into specific customer segments for insights We have found quite a few interesting insights that we can dive deeper into. We might want to target customer segments that contribute the most to sales to retain them or further increase sales. Let's look at Mainstream - young singles/couples. For instance, let's find out if they tend to buy a particular brand of chips.

```{r fig.align = "center"}
#### Deep dive into Mainstream, young singles/couples
set1 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER == "Mainstream",]
set2 <- data[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES") & PREMIUM_CUSTOMER != "Mainstream", ]
#Brand affinity 

quantity_set1 <- sum(set1$PROD_QTY)
quantity_set2 <- sum(set2$PROD_QTY)

target_segment <- set1 %>%
  group_by(BRAND) %>%
  summarise(target_segment = sum(PROD_QTY)/quantity_set1)
print(target_segment)
other_segment <- set2 %>%
  group_by(BRAND) %>%
  summarise(other_segment = sum(PROD_QTY)/quantity_set2)
print(other_segment)

brand_proportions <- merge(target_segment,other_segment)
brand_proportions$affinityToBrand <- brand_proportions$target_segment/brand_proportions$other_segment

print(brand_proportions[order(brand_proportions$affinityToBrand),])
# Over to you! Work out of there are brands that these two customer segments prefer more than others. You could use a technique called affinity analysis or a-priori analysis (or any other method if you prefer)
```

We can see that : [INSIGHTS] Let's also find out if our target segment tends to buy larger packs of chips.

```{r fig.align = "center"}
#### Preferred pack size compared to the rest of the population

packsize_target_segment <- set1 %>%
  group_by(PACK_SIZE) %>%
  summarise(packsize_target_segment = sum(PROD_QTY)/quantity_set1)
print(packsize_target_segment)
packsize_other_segment <- set2 %>%
  group_by(PACK_SIZE) %>%
  summarise(packsize_other_segment = sum(PROD_QTY)/quantity_set2)
print(packsize_other_segment)

pack_proportions <- merge(packsize_target_segment,packsize_other_segment)
pack_proportions$affinityToPack <- pack_proportions$packsize_target_segment/pack_proportions$packsize_other_segment

print(pack_proportions)
data[PACK_SIZE == 270, unique(PROD_NAME)]
data[PACK_SIZE == 330, unique(PROD_NAME)]

```

The mainstream young single/couple are more likely to buy 270 and 330g package which account for 30% and 37% more than the other pack size of chips. Investigating the brand name of them then we found that Twisties Cheese is the favourite 270g pack. The 330g packs have several options: Doritos Cheese, Smiths Crinkle Original and Smiths Crinkle Chips Salts and Vinegar flavour.

## [INSIGHTS]

It is clear that Mainstream young singles and couples are likely to spend more money on chips. The retiree shoppers also spend on it more than the other group of people.

They prefer Tyrells, Twisties, Kettle, Tostitos more than the other brands. This result could help to focus on those branding more than the others since they are the most preferable brands of the main customer segmentation.

The pack size of 270g and 330g also need to consider to focus because young single and couple prefer them more than the other. However, the Smiths for 2 flavor: "Original" and "Salts & Vinegar" are their favourite, it maybe because of the convenient pack size for them.
